What Happened to The Birth Rate?

### 2022-02-28 outputs/paper/paper.Rmd

---
title: "What Happened to The Birth Rate?"
subtitle: "Impact to Birth Rate Brought By The Great Recession and The LGBT Movement"
author: 
  - Lantao Cui
  - Huakun Shen
  - Sijian Wu
thanks: "Replication avaliable at https://doi.org/10.48152/ssrp-mbxx-jm47. Code and data avaliable at https://github.com/cuilantao/Investigation-of-falling-US-Birth-Rates-since-the-great-recession"         
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "This paper is based on Kearney's paper \"The Puzzle of Falling US Birth Rates since the Great Recession\". ^[Original paper available at: https://www.aeaweb.org/articles?id=10.1257/app.2.2.95] [@cite_original] During the Great Recession, the US birth rate varied between a tight range of roughly 65 to 70 births per 1000 women between the age of 15 to 44. However, after the Great Recession, the birth rate started to decline significantly, it dropped by nearly 20 percent in 13 years. Maintaining the fertility rate to be around 2.1 (replacement level) is important because low fertility rate could lead to aging of population and economic stagnation of economy. In this paper, we expand our research on multiple countries and found out that not only the US was affected by the recession, but the whole World in the context of birth rate, and the most affected countries are mostly developing countries.
We also treat the rise in the LGBT community as a potential cause for the decline in birth rate and examine its effect, and found that the growth of LGBT population may have a cohort effect and lead caused a negative impact on the birth rate."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
header-includes:
  - \usepackage{subfig}
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

@@

Another factor other than policy or economic factors that might have affected the fertility rate is cohort effect, more specifically, in this paper we focus on the rise of the LGBT community. The LGBT activity started in the 1920s and had a huge progress after the 2000s[@same_sex_marriage_states_2022]. A lot of laws that prohibited homosexual relationships were lifted and homosexual couples can get married legally and adopt kids, there is also propaganda that encourages homosexuality.

@@

The data used to analyze the impact to birth rate brought by LGBT population growth comes from the School of Law, UCLA[@will_institute]. The analysis was based on the data collected for the Gallup Daily Tracking survey.

@@

```{r, fig-past-40-years, fig.cap="US Birth Rate in the past 40 years", echo=FALSE, warning=FALSE, message = FALSE, fig.width=10, fig.height=4}
US_birth_date_from_1985_to_2020 <- read_csv(here::here("inputs/data/Birthrate.csv"))

ggplot(data=US_birth_date_from_1985_to_2020, aes(x=`year`, y=`brate_all`, group=1)) +
  geom_line()+
  geom_point() +
  ylim(50, 80) +
  xlab("Year") + ylab("Births per 1,000 women age 15-44")

@@

```{r cohort-plot, echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.cap="Children Ever Born by Mother Age and Birth Cohort", fig.height=4, fig.width=8}
age_comp_pop <- read_dta(here::here("inputs/data/agecomp-seer.dta"))
age_comp_pop <- subset(age_comp_pop, select = -c(fem1519, fem2034, fem3544, stname))
age_comp_pop_grouped <- group_by(age_comp_pop, year) %>% summarise(across(everything(), sum))
age_comp_pop_grouped_start_with_fem <-
  select(age_comp_pop_grouped, starts_with("fem"))
fem_col_name <- colnames(age_comp_pop_grouped_start_with_fem)
age_comp_pop_long <- gather(age_comp_pop_grouped, mage, fem, fem_col_name)
age_comp_pop_long$mage <- gsub("^.{0,3}", "", age_comp_pop_long$mage)
age_comp_pop_long$mage <- as.numeric(as.character(age_comp_pop_long$mage))
age_comp_pop_long$cohort <- with(age_comp_pop_long, year - mage)
age_comp_pop_long$cohort2 <- ''

age_comp_pop_long$cohort2[age_comp_pop_long$cohort >= 1968 & age_comp_pop_long$cohort <= 1972] <- 1
age_comp_pop_long$cohort2[age_comp_pop_long$cohort >= 1973 & age_comp_pop_long$cohort <= 1977] <- 2
age_comp_pop_long$cohort2[age_comp_pop_long$cohort >= 1978 & age_comp_pop_long$cohort <= 1982] <- 3
age_comp_pop_long$cohort2[age_comp_pop_long$cohort >= 1983 & age_comp_pop_long$cohort <= 1987] <- 4
age_comp_pop_long$cohort2[age_comp_pop_long$cohort >= 1988 & age_comp_pop_long$cohort <= 1992] <- 5
age_comp_pop_long$cohort2[age_comp_pop_long$cohort >= 1993 & age_comp_pop_long$cohort <= 1997] <- 6
age_comp_pop_long <- subset(age_comp_pop_long, cohort2 != '')

names(age_comp_pop_long)[names(age_comp_pop_long) == 'fem'] <- 'pop'
age_comp_pop_long_group <- group_by(age_comp_pop_long, cohort2, mage) %>% summarise(pop_sum = sum(pop))
cohort_analysis <- read_dta(here::here("inputs/data/nchs_cohort_analysis.dta"))
merged_data <- merge(cohort_analysis, age_comp_pop_long_group, by = c("cohort2", "mage"))
merged_data$brate <- with(merged_data, numbirth/pop_sum*1000)
data_with_only_cohort_and_brate <- subset(merged_data, select = c(cohort2, brate, mage))
data_with_only_cohort_and_brate <- data_with_only_cohort_and_brate[order(data_with_only_cohort_and_brate$cohort2, data_with_only_cohort_and_brate$mage),]
data_with_only_cohort_and_brate$brate <- ave(data_with_only_cohort_and_brate$brate, data_with_only_cohort_and_brate$cohort2, FUN=cumsum)
data_with_only_cohort_and_brate_grouped <- group_by(data_with_only_cohort_and_brate, cohort2, mage) %>% summarise(cum_brate = sum(brate) / 1000)
names(data_with_only_cohort_and_brate_grouped)[names(data_with_only_cohort_and_brate_grouped) == 'cohort2'] <- 'cohort'
final_data_wide_format <- spread(data_with_only_cohort_and_brate_grouped, key = cohort, value = cum_brate,  sep = '')

data_with_cohort1 <- drop_na(final_data_wide_format, cohort1)
data_with_cohort2 <- drop_na(final_data_wide_format, cohort2)
data_with_cohort3 <- drop_na(final_data_wide_format, cohort3)
data_with_cohort4 <- drop_na(final_data_wide_format, cohort4)
data_with_cohort5 <- drop_na(final_data_wide_format, cohort5)
data_with_cohort6 <- drop_na(final_data_wide_format, cohort6)

ggplot() +
  geom_line(data = data_with_cohort1, aes(x = mage, y = cohort1, color = "1968-1972"))+
  geom_line(data = data_with_cohort2, aes(x = mage, y = cohort2, color = "1973–1977")) +
  geom_line(data = data_with_cohort3, aes(x = mage, y = cohort3, color = "1977–1982")) +
  geom_line(data = data_with_cohort4, aes(x = mage, y = cohort4, color = "1983–1987")) +
  geom_line(data = data_with_cohort5, aes(x = mage, y = cohort5, color = "1988–1992")) +
  geom_line(data = data_with_cohort6, aes(x = mage, y = cohort6, color = "1993–1997")) +
  scale_color_manual(name = "Birth Cohorts", 
                     values = c("1968-1972" = "green", 
                                "1973–1977" = "blue", 
                                "1977–1982" = "orange", 
                                "1983–1987" = "gray",
                                "1988–1992" = "yellow4",
                                "1993–1997" = "lightblue"
                                )) + theme(legend.position = c(0.9, 0.5)) +
  xlab("Mother's Age") + ylab("Children ever born") + 
  ggtitle("Children Ever Born by Mother’s Age and Birth Cohort")
```

@@

```{r fig-parity, fig.cap="Parity (ages 15-44)", echo=FALSE, fig.height=4, fig.height=3}
birth_pop <- read_dta(here::here("inputs/data/nchs_births_pop_1990_2019.dta"))
age_race_pop <- read_dta(here::here("inputs/data/age_race_comp_seer.dta"))

merged_data <- merge(birth_pop, age_race_pop, by = c("stname", "year"))

# Define new column
merged_data$pop2044 <- with(merged_data, `pop2024.y` + `pop2534.y` + `pop3544.y`)

merged_data_grouped_by <- group_by(merged_data, year) %>% summarise(fb_sum = sum(numbirth_firstbirth), sb_sum = sum(numbirth_secondbirth), tb_sum = sum(numbirth_thirdbirth), fbp_sum = sum(numbirth_fourthbirth), pop_sum = sum(pop1544))

merged_data_grouped_by$brith_rate_fb <- with(merged_data_grouped_by, fb_sum / `pop_sum` * 1000)
merged_data_grouped_by$brith_rate_sb <- with(merged_data_grouped_by, sb_sum / `pop_sum` * 1000)
merged_data_grouped_by$brith_rate_tb <- with(merged_data_grouped_by, tb_sum / `pop_sum` * 1000)
merged_data_grouped_by$brith_rate_fbp <- with(merged_data_grouped_by, fbp_sum / `pop_sum` * 1000)

ggplot(merged_data_grouped_by, aes(x=year)) + 
 geom_line(aes(y = brith_rate_fb, color = "First Birth")) + 
  geom_line(aes(y = brith_rate_sb, color="Second Birth")) +
  geom_line(aes(y = brith_rate_tb, color="Third Birth")) +
  geom_line(aes(y = brith_rate_fbp, color="Fourth Birth")) +
  scale_color_manual(name = "Parity (ages 15-44)", 
                     values = c("First Birth" = "steelblue", 
                                "Second Birth" = "orange", 
                                "Third Birth" = "gray", 
                                "Fourth Birth" = "yellow")) + 
  xlab("Year") + ylab("Birth Rate (births per 1000 women)") + ggtitle("Parity (ages 15-44)")
```

>>>>

---
title: "What Happened to The Birth Rate?"
subtitle: "Impact to Birth Rate Brought By The Great Recession and The LGBT Movement"
author: 
  - Lantao Cui
  - Huakun Shen
  - Sijian Wu
thanks: "Replication avaliable at https://doi.org/10.48152/ssrp-mbxx-jm47. Code and data avaliable at https://github.com/cuilantao/Investigation-of-falling-US-Birth-Rates-since-the-great-recession"         
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "This paper is based on Kearney's paper \"The Puzzle of Falling US Birth Rates since the Great Recession\". ^[Original paper available at: https://www.aeaweb.org/articles?id=10.1257/app.2.2.95] [@cite_original] During the Great Recession, the US birth rate varied between a tight range of roughly 65 to 70 births per 1000 women between the age of 15 to 44. However, after the Great Recession, the birth rate started to decline significantly, it dropped by nearly 20 percent in 13 years. Maintaining the fertility rate to be around 2.1 (replacement level) is important because low fertility rate could lead to aging of population and economic stagnation of economy. In this paper, we expanded our research on multiple countries and found out that not only the US was affected by the recession, but the whole World in the context of birth rate, and the most affected countries are mostly developing countries. We also treat the rise in the LGBT community as a potential cause for the decline in birth rate and examined its effect, and found that the growth of LGBT population may have a cohort effect and lead caused a negative impact on the birth rate."
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
header-includes:
  - \usepackage{subfig}
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

@@

Another factor other than policy or economic factors that might have affected the fertility rate is cohort effect, more specifically, in this paper we focus on the rise of the LGBT community. The LGBT activity started in the 1920s and had a huge progress after the 2000s[@same_sex_marriage_states_2022]. A lot of laws that prohibited homosexual relationships were lifted and homosexual couples can get married legally and adopt kids, there is also propaganda that encourages homosexuality. Birth rates and LGBT was found to be highly correlated, the growth of LGBT population could lead to lower birth rate and the growth of LGBT population is increasing rapidly in the younger generations.

@@

The data used to analyze the impact to birth rate brought by LGBT population growth comes from the School of Law, UCLA[@will_institute]. The analysis was based on the data collected for the Gallup Daily Tracking survey. Respondents were obtained through list-assisted random digit dial (70% cell phone, 30% landline).

@@

```{r, fig-past-40-years, fig.cap="US Birth Rate in the past 40 years", echo=FALSE, warning=FALSE, message = FALSE, fig.width=10, fig.height=4}
# import the csv
US_birth_date_from_1985_to_2020 <- read_csv(here::here("inputs/data/Birthrate.csv"))

# Plot the graph
ggplot(data=US_birth_date_from_1985_to_2020, aes(x=`year`, y=`brate_all`, group=1)) +
  geom_line()+
  geom_point() +
  ylim(50, 80) +
  xlab("Year") + ylab("Births per 1,000 women age 15-44")
```

@@

```{r cohort-plot, echo=FALSE, results='hide', message=FALSE, warning=FALSE, fig.cap="Children Ever Born by Mother Age and Birth Cohort", fig.height=4, fig.width=8}
# Read the dta
age_comp_pop <- read_dta(here::here("inputs/data/agecomp-seer.dta"))
# Filter out unwanted columns
age_comp_pop <- subset(age_comp_pop, select = -c(fem1519, fem2034, fem3544, stname))
# group by and find the sum of all columns
age_comp_pop_grouped <- group_by(age_comp_pop, year) %>% summarise(across(everything(), sum))
# generate vector of column names that start with fem
age_comp_pop_grouped_start_with_fem <-
  select(age_comp_pop_grouped, starts_with("fem"))
fem_col_name <- colnames(age_comp_pop_grouped_start_with_fem)
# transform wide to long
age_comp_pop_long <- gather(age_comp_pop_grouped, mage, fem, fem_col_name)
# generate new columns
age_comp_pop_long$mage <- gsub("^.{0,3}", "", age_comp_pop_long$mage)
# transform it into a numeric value
age_comp_pop_long$mage <- as.numeric(as.character(age_comp_pop_long$mage))
age_comp_pop_long$cohort <- with(age_comp_pop_long, year - mage)
age_comp_pop_long$cohort2 <- ''
# assign different numbers based on years
age_comp_pop_long$cohort2[age_comp_pop_long$cohort >= 1968 & age_comp_pop_long$cohort <= 1972] <- 1
age_comp_pop_long$cohort2[age_comp_pop_long$cohort >= 1973 & age_comp_pop_long$cohort <= 1977] <- 2
age_comp_pop_long$cohort2[age_comp_pop_long$cohort >= 1978 & age_comp_pop_long$cohort <= 1982] <- 3
age_comp_pop_long$cohort2[age_comp_pop_long$cohort >= 1983 & age_comp_pop_long$cohort <= 1987] <- 4
age_comp_pop_long$cohort2[age_comp_pop_long$cohort >= 1988 & age_comp_pop_long$cohort <= 1992] <- 5
age_comp_pop_long$cohort2[age_comp_pop_long$cohort >= 1993 & age_comp_pop_long$cohort <= 1997] <- 6
age_comp_pop_long <- subset(age_comp_pop_long, cohort2 != '')

names(age_comp_pop_long)[names(age_comp_pop_long) == 'fem'] <- 'pop'
age_comp_pop_long_group <- group_by(age_comp_pop_long, cohort2, mage) %>% summarise(pop_sum = sum(pop))
# read data files
cohort_analysis <- read_dta(here::here("inputs/data/nchs_cohort_analysis.dta"))
# merge two file together
merged_data <- merge(cohort_analysis, age_comp_pop_long_group, by = c("cohort2", "mage"))
# define new columns
merged_data$brate <- with(merged_data, numbirth/pop_sum*1000)
data_with_only_cohort_and_brate <- subset(merged_data, select = c(cohort2, brate, mage))
data_with_only_cohort_and_brate <- data_with_only_cohort_and_brate[order(data_with_only_cohort_and_brate$cohort2, data_with_only_cohort_and_brate$mage),]
# accmulate all the values
data_with_only_cohort_and_brate$brate <- ave(data_with_only_cohort_and_brate$brate, data_with_only_cohort_and_brate$cohort2, FUN=cumsum)
data_with_only_cohort_and_brate_grouped <- group_by(data_with_only_cohort_and_brate, cohort2, mage) %>% summarise(cum_brate = sum(brate) / 1000)
names(data_with_only_cohort_and_brate_grouped)[names(data_with_only_cohort_and_brate_grouped) == 'cohort2'] <- 'cohort'
final_data_wide_format <- spread(data_with_only_cohort_and_brate_grouped, key = cohort, value = cum_brate,  sep = '')
# drop all na value
data_with_cohort1 <- drop_na(final_data_wide_format, cohort1)
data_with_cohort2 <- drop_na(final_data_wide_format, cohort2)
data_with_cohort3 <- drop_na(final_data_wide_format, cohort3)
data_with_cohort4 <- drop_na(final_data_wide_format, cohort4)
data_with_cohort5 <- drop_na(final_data_wide_format, cohort5)
data_with_cohort6 <- drop_na(final_data_wide_format, cohort6)
# draw the plot
ggplot() +
  geom_line(data = data_with_cohort1, aes(x = mage, y = cohort1, color = "1968-1972"))+
  geom_line(data = data_with_cohort2, aes(x = mage, y = cohort2, color = "1973–1977")) +
  geom_line(data = data_with_cohort3, aes(x = mage, y = cohort3, color = "1977–1982")) +
  geom_line(data = data_with_cohort4, aes(x = mage, y = cohort4, color = "1983–1987")) +
  geom_line(data = data_with_cohort5, aes(x = mage, y = cohort5, color = "1988–1992")) +
  geom_line(data = data_with_cohort6, aes(x = mage, y = cohort6, color = "1993–1997")) +
  scale_color_manual(name = "Birth Cohorts", 
                     values = c("1968-1972" = "green", 
                                "1973–1977" = "blue", 
                                "1977–1982" = "orange", 
                                "1983–1987" = "gray",
                                "1988–1992" = "yellow4",
                                "1993–1997" = "lightblue"
                                )) + theme(legend.position = c(0.9, 0.5)) +
  xlab("Mother's Age") + ylab("Children ever born") + 
  ggtitle("Children Ever Born by Mother’s Age and Birth Cohort")
```

@@

```{r fig-parity, fig.cap="Parity (ages 15-44)", echo=FALSE, fig.height=4, fig.height=3}
# Import the dta
birth_pop <- read_dta(here::here("inputs/data/nchs_births_pop_1990_2019.dta"))
age_race_pop <- read_dta(here::here("inputs/data/age_race_comp_seer.dta"))
merged_data <- merge(birth_pop, age_race_pop, by = c("stname", "year"))

# Define new column
merged_data$pop2044 <- with(merged_data, `pop2024.y` + `pop2534.y` + `pop3544.y`)
# use a group by to get the sum
merged_data_grouped_by <- group_by(merged_data, year) %>% summarise(fb_sum = sum(numbirth_firstbirth), sb_sum = sum(numbirth_secondbirth), tb_sum = sum(numbirth_thirdbirth), fbp_sum = sum(numbirth_fourthbirth), pop_sum = sum(pop1544))
#define new columns based on existing columns
merged_data_grouped_by$brith_rate_fb <- with(merged_data_grouped_by, fb_sum / `pop_sum` * 1000)
merged_data_grouped_by$brith_rate_sb <- with(merged_data_grouped_by, sb_sum / `pop_sum` * 1000)
merged_data_grouped_by$brith_rate_tb <- with(merged_data_grouped_by, tb_sum / `pop_sum` * 1000)
merged_data_grouped_by$brith_rate_fbp <- with(merged_data_grouped_by, fbp_sum / `pop_sum` * 1000)
# merge the plot and add legends
ggplot(merged_data_grouped_by, aes(x=year)) + 
 geom_line(aes(y = brith_rate_fb, color = "First Birth")) + 
  geom_line(aes(y = brith_rate_sb, color="Second Birth")) +
  geom_line(aes(y = brith_rate_tb, color="Third Birth")) +
  geom_line(aes(y = brith_rate_fbp, color="Fourth Birth")) +
  scale_color_manual(name = "Parity (ages 15-44)", 
                     values = c("First Birth" = "steelblue", 
                                "Second Birth" = "orange", 
                                "Third Birth" = "gray", 
                                "Fourth Birth" = "yellow")) + 
  xlab("Year") + ylab("Birth Rate (births per 1000 women)") + ggtitle("Parity (ages 15-44)")
```
